Quantitative Benchmarking for Automatically Generated Conjectures

### 2018-04-30 TEBenchmarkPaper/article.tex
TEB:More fixes

% FIXME: Maybe rethink "analysis" bit; maybe add summary and comparison
% references (although maybe not as full blown steps)
\begin{figure}
  \tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm,
                           minimum height=1cm,text centered, draw=black]

@@

\subsection{Tons of Inductive Problems}

@@

% FIXME: The designers must decide on, not the system
Fundamentally, any mathematical reasoning system must decide on, and formalise,
what counts as ``the good'' in mathematics.  Obvious metrics such as ``true'' or
``provable'' include trivial tautologies, while at the same time failing to
capture the ``almost true'', which can be a valuable trigger for theory change,
as demonstrated by Lakatos in his case studies of mathematical
development~\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly
proposed metric. Neuro-scientists such as Zeki \etal{} have attempted to shed
light on this by testing whether mathematicians' experiences of abstract beauty
correlates with the same brain activity as experiences of sensory
beauty~\cite{10.3389/fnhum.2014.00068}. Metrics such as those in
table~\ref{table:colton}, from computer scientists like Colton, are based
largely on ``intuition'', plausibility arguments about why a metric would be
important, and previous use of such metrics (in the case of
``surprisingingness'', from a single quote from a mathematician).
Opinions from mathematicians themselves include Gowers' suggestion that we can
identify features which are commonly associated with good
proofs~\cite{gowers2000two}, Erdos's famous idea of ``The Book'' (a theoretical
book containing the most elegant proof of every theorem, brought to life
in~\cite{aigner2010proofs}) as well as McCasland's evaluation of the
interestingness of MATHsAiD's output~\cite{roy} (of which he was the main system
developer).

@@

One notable weakness is that our methodology only makes sense in domains where
existing human knowledge surpasses that discovered by the machine. Any
\emph{truly novel} insights discovered during testing will not, by definition,
appear in any existing corpus, and we would in fact \emph{penalise} tools for
such output. We do not believe this to be a realistic problem for the time
being, although we would be delighted if proved wrong. % FIXME: Is there a solution? Mention that this is common to most tests

Another practical limitation of our benchmarking approach is that it only
applies to tools which act in ``batch mode'', i.e. those which choose to halt
after emitting some output. Whilst all of the systems we have encountered are of
this form, some (such as IsaCoSy, QuickSpec 2 and Speculate) could conceivably
be run without a depth limit, and form part of the ``scientist assistant'' role
envisaged by Lenat (or McCarthy's earlier ``advice
taker''~\cite{McCarthy_Programs59}). Analogous benchmarks could be designed for
such long-running, potentially interactive programs, but that is beyond the
scope of this project.

@@

% FIXME: Reword into a short intro, e.g. "There is a long history of automating
% mathematics, going back to... A recurring theme is separation between the
% efforts of pure mathematicians and those wanting practical engineering
% applications....
The first automated mathematical tool, the mechanical calculator
(known as the Pascaline), was an adding machine that could perform
additions and subtractions directly and multiplication and divisions
by repetitions, and was conceived by Pascal in 1642 while reorganising
tax revenues~\cite{d'ocagne}. Subsequent early tools include
M\"uller's universal calculating machine in 1784, which he built for
the purpose of calculating and printing numerical tables: he invented
this when he had to check and recalculate some tables relating to the
volumes of trees~\cite[p. 65]{lindgren}. Thirty seven years later,
Babbage invented his famous difference engine: an automatic,
mechanical calculator designed to tabulate polynomial functions. This
was inspired by a flawed table of logarithms and the idea that
machines would be quicker and more reliable~\cite{bowden}. It is
interesting to note background and motivation: while Pascal and
Babbage were mathematicians, with an interest in engineering, M\"uller
was an engineer with an interest in mathematical knowledge. All three
tools were conceived as an aid to mathematicians, as well as
scientists, accountants and surveyors.

Differences in background and motivation continue to be relevant
today. While the majority of work in automating mathematics has been
in symbolic manipulation and theorem proving, we are concerned here
with other aspects of mathematics, including the construction of conjectures,
(counter\nobreakdash-)examples and theorems. These tasks, along with the
formation of new definitions and concepts, are varyingly known as
``Automated Theory Formation''~\cite{lenat:77,colton:book},
``Mathematical Theory Exploration''~\cite{buchberger:06} (also sometimes
prefaced with ``Computer-Aided'', ``Automated'' or ``Algorithm-Supported''),
``Automated Mathematical Discovery''~\cite{epstein:91,colton:interestingness,esarm2008},
``Concept Formation in Discovery Systems''~\cite{haase}, and
``Automated Theorem Discovery''~\cite{roy}. Such a plethora of terminology can
be unhelpful and can mask similarities between the different fields. In
particular, the twin strands of Automated Theory Formation and
Automated Mathematical Theory Exploration seem to be developing
somewhat independently without a clear differentiating
methodology. Below we discuss commonalities and differences between
the two schools of thought.

\subsection{Automated Theory Formation}
\label{section:atf}

% FIXME: Cut this down a lot. Currently it's not even an ATF section, it's just
% an AM section. We should maybe remove its status as a section, and instead
% just remark on the influence of AM on the field

% 1. history of terminology
Automated theory formation derives its terminology from psychology in which the
term ``concept formation'' is used (see, for example,~\cite{bruner:67}) to
describe the search for features which differentiate exemplars from
non-exemplars of various categories. Lenat used this term in his 1977 paper:
{\em Automated Theory Formation in Mathematics}~\cite{lenat:77}.

% 4. example tools - AM, HR, ...
When Lenat built the AM system~\cite{lenat:77}, there were systems
which could define new concepts for investigation, such as those
described in~\cite{winston}, and systems which could discover
relationships among known concepts, such as Meta-Dendral~\cite{buchanan:75}. No
system could perform both of these tasks: Lenat saw this as the next step. AM
was designed to both construct new concepts and conjecture relationships between
them; fully automating the cycle of discovery in mathematics. Lenat describes
this as follows:

\begin{quote}
``What we are describing is a computer program which
defines new concepts, investigates them, notices
regularities in the data about them, and conjectures
relationships between them. This new information is used
by the program to evaluate the newly-defined concepts,
concentrate upon the most interesting ones, and iterate the
entire process.''~\cite[p. 834]{lenat:77}
\end{quote}

AM was a rule-based system which used a frame-like scheme to represent
its knowledge, enlarged its knowledge base via a collection of
heuristic rules, and controlled the firing of these rules via an
agenda mechanism. Lenat chose elementary arithmetic as the development
domain because he could use personal introspection for the heuristics
for constructing and evaluating concepts. Given the age of this
discipline, Lenat thought it unlikely that AM would make significant
discoveries, although he did cite its ``ultimate achievements'' as the
concepts and conjectures it discovered (or could have discovered). He
suggested various criteria by which his system could be evaluated,
many of which focused on an exploration of the techniques. For
instance, he considered generality (running AM in new domains) and how
finely-tuned various aspects of the program are (the agenda, the
interaction of the heuristics, etc). %most of which were qualitative.
Lenat saw his system and future developments in this field as having
implications for mathematics itself (finding results of significance),
for automating mathematics research (developing AI techniques), and
for designing ``scientist assistant'' programs (aids for
mathematicians). This shows a broad spread of motivation. Despite the
seeming success of the AM system, it is one of the most criticised
pieces of AI research. In their case study in methodology, Ritchie and
Hanna analysed Lenat's written work on AM and found that there was a
large discrepancy between his theoretical claims and the implemented
program~\cite{partridge}. For instance, Lenat made claims about how AM
invented natural numbers from sets, whereas it used one heuristic
which was specifically written in order to make this connection (and
not used in any other context). Another problem was that the processes
were sometimes under-explained. For an argument of why many of the
claims made by Lenat about AM were false, see chapter 13 of~\cite{colton:book}.

\subsection{Mathematical Theory Exploration}
\label{section:mte}

The phrase Mathematical Theory Exploration (MTE) is a recent term % FIXME: When?
which seems to have originated with Buchberger and colleagues (see, % FIXME: informal
for example,~\cite{buchberger}). His motivation is to support
mathematicians during their exploration of mathematical theories. This
support is intended to be for the straightforward reasoning, which he
argues, covers most mathematical thought, rather than the ingenious
points, which he leaves for human mathematicians. Buchberger's long
term goal is to provide routine tools for the exploration activity of
working mathematicians, to support the invention and structured
build-up of mathematical knowledge. The Theorema project aims at
prototyping features of a system for such a purpose. These features
include Integration of the Functionality of Current Mathematical
Systems (retention of the full power of current numerics and computer
algebra systems, as well as enabling the user to add their own
algorithms to the system); Attractive Syntax (input and output is
readable and presented attractively, and can be personalised by the
user); and Structured Mathematical Knowledge Bases (tools are provided
for building and using large mathematical knowledge
libraries). Buchberger has evaluated the potential of this strategy by
illustrating the automated synthesis of his own Gr\"obner bases
algorithm~\cite{buchberger:04}.

Recent systems developed in this area include IsaScheme~\cite{MontanoRivas2011},
IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation},
QuickSpec~\cite{QuickSpec} and Speculate~\cite{braquehais2017speculate}
(included in table~\ref{table:colton}) as well as MATHsAiD~\cite{roy}.

The goal of the MATHsAiD (Mechanically Ascertaining Theorems from Hypotheses,
Axioms and Definitions) project is to build a tool which takes in a set of
axioms, concept definitions and a logic and applies its inference rules to
reason from the axioms to theorems. The motivation is to produce a tool which
will help mathematicians to explore the consequences of a set of axioms or a
particular concept definition. One of the main challenges of the project has
been to automatically evaluate interestingness: to distinguish important
theorems from results which, although they follow from a set of axioms, are of
little mathematical interest.

% IsaScheme FIXME: Maybe strip down this description? We could use prose instead,
% like "schemes capture common patterns such as iteration, which can be
% instantiated with an addition function to produce multiplication, or with
% multiplication to produce exponentiation.
Monta{\~n}o-Rivas \etal{} have implemented a scheme-based approach to MTE in their
IsaScheme system~\cite{MontanoRivas2011}. Schemes are higher-order formulae
which can be used to generate new concepts and conjectures; variables within the
scheme are instantiated automatically and this drives the invention process.
For instance, in the theory of natural numbers, given the concepts of zero
($\Zero$), successor ($\Succ$) and addition ($\plus$), IsaScheme can use the
following scheme to invent the concept of multiplication:

\[
\text{def-scheme}(g,h,i,j) \equiv\\
  \exists f. \forall x y. \left\{
  \begin{array}{l l}
    f(g,y) = h (y)  \: \text{and}  & \\
    f(i(x),y) = j(y,f(x,y)) & \\
  \end{array} \right.
\]

where the existentially quantified variable $f$ stands for the new
function to be defined in terms of the variables $g$, $h$, $i$ and $j$. Within
the theory of natural numbers, IsaScheme instantiates this scheme with
$\sigma_1 = \{g \mapsto \Zero, h \mapsto (\lambda x.\Zero), i \mapsto \Succ, j
\mapsto \plus\}$. In this example, $f \mapsto \mult$ (multiplication), since:

\begin{align*}
\mult(\Zero, y)  &= \Zero \\
\mult(\Succ(x), y) &= \plus(y, \mult(x, y))
\end{align*}

The new multiplication function $f$ can itself be used to instantiate
variables in the same scheme, resulting in the invention of the
exponentiation concept (these examples are taken from~\cite{MontanoRivas2011}).

% FIXME: CITE
QuickSpec version 2 replaces this two-step process with a single, iterative
algorithm similar to that of IsaCoSy. Generated conjectures are fed into a
Knuth-Bendix completion algorithm to form a corresponding set of rewrite rules.
As expressions are enumerated, they are simplified using these rules and
discarded if equal to a known expression. If not, QuickCheck tests whether the
new expression can be distinguished from the known expressions through random
testing: those which can are added to the set of known expressions. Those which
cannot be distinguished are conjectured to be equal, and the rewrite rules are
updated.

%Speculate
Speculate is a Haskell tool which operates in a similar way to QuickSpec
version 2. In addition to equations, Speculate uses the laws of total orders and
Boolean algebra, along with testing (via LeanCheck), to conjecture inequalities
and conditional relations between expressions, such as (for
$x, y \in \mathbb{Z}$):

\begin{equation*}
  \begin{aligned}
    x + y \leq \lvert x \rvert + \lvert y \rvert \\
    x \leq 0 \rightarrow x + \lvert x \rvert = 0
  \end{aligned}
\end{equation*}

\subsection{Commonalities and Differences between MTE and ATF}

% FIXME: Condense
ATF and MTE share many goals and would benefit from a
closer alignment: both fields are highly specialised and by joining
forces (while being explicit about differences) the techniques being
developed would have greater impact. This closer alignment is already
beginning to take place. The main commonalities between MTE and ATF
are that both are interested in the same aspects of mathematical
thinking, aiming to automatically construct and evaluate elements of a
mathematical theory, including concepts, conjectures, theorems, axioms
and examples. Both strands contrast themselves with Automated Theorem
Proving. The main historical differences between these two approaches seem to be
that MTE was initiated by mathematicians as an attempt to support existing
mathematical research, perhaps with user interaction; while ATF is a product of
AI researchers trying to extend AI techniques, with full automation and often
applications to non-mathematical domains.

The different backgrounds of the people who named the fields
``automated theory formation'' (Lenat) and ``mathematical theory
exploration'' (Buchberger) perhaps reflects different metaphysical
perspectives on invention (in which case new mathematical ideas are
being {\em formed}, created or produced) and discovery (in which case
abstract mathematical objects exist independently of us and are {\em
  explored} or investigated) in mathematics.\footnote{There is a vast
  literature on the Platonic and the constructivist view of
  mathematics: we shall not venture down this path here (interested
  readers are referred to~\cite{hersh:97,shapiro}). Instead, for now,
  we make the pragmatic assumption that while there may well be
  different cognitive processes involved in invention than those
  involved in discovery, the fields of ATF and MTE are not currently
  concerned with this level of detail and so the philosophical
  distinction is not relevant for our purposes.} As can be seen above,
some of these historical differences are disappearing, and recent
implementations are bridging the methodological gap between ATF and MTE. This is
particularly true in terms of fully-automatic/user-interactive and the
domain-specific/general aspects.

>>>>

\begin{figure}
  \tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm,
                           minimum height=1cm,text centered, draw=black]

@@

\subsection{Tons of Inductive Problems}
\label{sec:tip}

@@

Fundamentally, when designing any mathematical reasoning system we must decide
on, and formalise, what counts as ``the good'' in mathematics.  Obvious metrics
such as ``true'' or ``provable'' include trivial tautologies, while at the same
time failing to capture the ``almost true'', which can be a valuable trigger for
theory change, as demonstrated by Lakatos in his case studies of mathematical
development~\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly
proposed metric. Neuro-scientists such as Zeki \etal{} have attempted to shed
light on this by testing whether mathematicians' experiences of abstract beauty
correlates with the same brain activity as experiences of sensory
beauty~\cite{10.3389/fnhum.2014.00068}. Qualities from computer scientists like
Colton (such as those in~\cite{colton2000notion}) are based largely on
``intuition'', plausibility arguments about why a metric would be important, and
previous use of such metrics (in the case of ``surprisingingness'', from a
single quote from a mathematician). Opinions from mathematicians themselves
include Gowers' suggestion that we can identify features which are commonly
associated with good proofs~\cite{gowers2000two}, Erdos's famous idea of ``The
Book''~\cite{aigner2010proofs} as well as McCasland's personal evaluation of the
interestingness of MATHsAiD's output~\cite{roy} (of which he was the main system
developer).

@@

One notable weakness is that our methodology only makes sense in domains where
existing human knowledge surpasses that discovered by the machine. Any
\emph{truly novel} insights discovered during testing will not, by definition,
appear in any existing corpus, and we would in fact \emph{penalise} tools for
such output. We do not believe this to be a realistic problem for the time
being, although it does emphasise the continued importance of testing these
tools with real human users, rather than relying solely on artificial benchmarks
such as ours.

Another practical limitation of our benchmarking approach is that it only
applies to tools which act in ``batch mode'', i.e. those which choose to halt
after emitting some output. Whilst all of the systems we have encountered are of
this form, some (such as IsaCoSy, QuickSpec 2 and Speculate) could conceivably
be run without a depth limit, and form part of the ``scientist assistant'' role
which Lenat envisaged for AM, or McCarthy's earlier ``advice
taker''~\cite{McCarthy_Programs59}. Analogous benchmarks could be designed for
such long-running, potentially interactive programs, but that is beyond the
scope of this project.

@@

The automation of mathematical tasks has been pursued since at least the time of
mechanical calculators like the Pascaline~\cite{d'ocagne}. A recurring theme in
these efforts is the separation between those undertaken by mathematicians like
Pascal and Babbage~\cite{bowden}, and those of engineers such as
M\"uller~\cite[p. 65]{lindgren}. This pattern continues today, with the tasks we
are concerned with (automatically constructing and evaluating concepts,
conjectures, theorems, axioms, examples, etc.) being divided into two main
fields: Mathematical Theory Exploration (MTE)~\cite{buchberger:06} (also
sometimes prefaced with ``Computer-Aided'', ``Automated'' or
``Algorithm-Supported''), which is championed by mathematicians such as
Buchberger~\cite{buchberger}; and Automated Theory Formation
(ATF)~\cite{lenat:77,colton:book}, pursued by AI researchers including Lenat.
Other related terms include ``Automated Mathematical
Discovery''~\cite{epstein:91,colton:interestingness,esarm2008},
``Concept Formation in Discovery Systems''~\cite{haase}, and
``Automated Theorem Discovery''~\cite{roy}.

Such a plethora of terminology can mask similarities and shared goals between
these fields. Even notable historical differences, such as the emphasis of MTE
on user-interaction and mathematical domains, in contrast to the full automation
and more general applications targeted by ATF, are disappearing in recent
implementations.

An important historical implementation of ATF is Lenat's AM system. Unlike prior
work, such as Meta-Dendral~\cite{buchanan:75} and those described
in~\cite{winston}, AM aims to be a general purpose mathematical discovery
system, designed to both construct new concepts and conjecture relationships
between them. AM is a rule-based system which represents knowledge using a
frame-like scheme, enlarges its knowledge base via a collection of heuristic
rules, and controls the firing of these rules via an agenda
mechanism. Evaluation of AM considered generality (performance in new domains)
and how finely-tuned various aspects of the program are (the agenda, the
interaction of the heuristics, etc). Most of this evaluation was qualitative,
and has subsequently been criticised~\cite[chap.~13]{colton:book}. In their
case study in methodology, Ritchie and Hanna found a large discrepancy between
the theoretical claims made of AM and the implemented program~\cite{partridge};
for example, AM ``invented'' natural numbers from sets, but did so using a
heuristic specifically designed to make this connection.

The prototypical implementation of MTE is the Theorema system of Buchberger and
colleagues~\cite{buchberger,buchberger2016theorema}, which also places a strong
emphasis on user interface and output presentation. Theory exploration in the
Theorema system involves the user formalising their definitions in a consistent,
layered approach; such that reasoning algorithms can exploit this structure in
subsequent proofs, calculations, etc. The potential of this strategy was
evaluated by illustrating the automated synthesis of Buchberger's own Gr\"obner
bases algorithm~\cite{buchberger:04}.

A similar ``layering'' approach is found in the IsaScheme system of
Monta{\~n}o-Rivas \etal{}~\cite{MontanoRivas2011}, which is compared against
IsaCoSy and HipSpec is compared using precision/recall
in~\cite{claessen2013automating}. The name comes from its embedding in the
Isabelle proof assistant and its use of ``schemes'': higher-order formulae which
can be used to generate new concepts and conjectures. Variables within a scheme
are instantiated automatically and this drives the invention process. For
example, the concept of ``repetition'' can be encoded as a scheme, and
instantiated with existing encodings of zero, successor and addition to produce
a definition of multiplication. The same scheme can be instantiated with this
new multiplication function to produce exponentiation.

IsaCoSy and QuickSpec (the conjecture generation component of HipSpec) are
described in more detail in $\S$\ref{sec:existing-tools}, since these are the
tools we chose to evaluate and compare to demonstrate our benchmarking
methodology. QuickSpec has since evolved to version 2~\cite{smallbone2017quick},
which replaces the distinct enumeration and testing steps with a single,
iterative algorithm similar to that of IsaCoSy. Generated conjectures are fed
into a Knuth-Bendix completion algorithm to form a corresponding set of rewrite
rules. As expressions are enumerated, they are simplified using these rules and
discarded if equal to a known expression. If not, QuickCheck tests whether the
new expression can be distinguished from the known expressions through random
testing: those which can are added to the set of known expressions. Those which
cannot be distinguished are conjectured to be equal, and the rewrite rules are
updated.

QuickSpec has also inspired another MTE tool for Haskell called
Speculate~\cite{braquehais2017speculate}, which operates in a similar way but
also makes use of the laws of total orders and Boolean algebra to conjecture
\emph{in}equalities and conditional relations between expressions.

Another notable MTE implementation, distinct from those based in Isabelle and
Haskell, is the MATHsAiD project (Mechanically Ascertaining Theorems from
Hypotheses, Axioms and Definitions)~\cite{roy}. The project's goal is to help
mathematicians explore the consequences of a set of axioms or a particular
concept definition by deducing, via the chosen logic, a set of theorems which
they entail. Like other MTE approaches, one of its main challenges is to
automatically evaluate interestingness: to distinguish important theorems from
results which, although they follow from a set of axioms, are of little
mathematical interest. Evaluation of these theorems was performed by the
system's developer, which highlights the need for a general method of evaluation
and comparison for these tools.

Whilst there are many reasonably objective benchmarks for mathematical tasks
such as automated theorem proving, the precision/recall analysis shown
in~\cite{claessen2013automating}, and described further in $\S$\ref{sec:te}, is
the only quantitative comparison of these recent tools we have found in the
literature. Our work is essentially an extension of this approach, to a larger
and more diverse set of examples. The suitability of the TIP theorem proving
benchmark for our purposes, detailed in $\S$\ref{sec:tip}, is not coincidental,
since its developers are also those of the IsaCoSy and QuickSpec tools we have
tested. This goes some way to ensuring that our demonstration is a faithful
representation of the task these tools were designed for, whilst our independent
repurposing of the problem set reduces the risk that the benchmark is
tailor-made for these tools; or, conversely, that the tools over-fit to these
particular problems.
